The Power of Unbiased Recursive Partitioning:\\ A Unifying View of CTree, MOB, and GUIDE

### 2020-06-14 pkg/partykit/inst/guideliketest/draft/draft.Rnw
added explanations/comments based on reviews
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3017 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 4c2b1bea2e4ed05de981fa68b5503c7ea74db959

%\fixme{``Why CTree, GUIDE, and MOB?''}
%
Therefore, in this paper the focus will be put on that first step of tree algorithms,
i.e., the task of selecting the best split variable in a given (sub)sample. 
In particular, the approach from the GUIDE algorithm (which essentially encompasses QUEST)
is compared to the one of the CTree and MOB algorithms by investigating the building blocks of 
their testing strategies in which they differ:
(1)~Variation of the goodness-of-fit measure for the response: using residuals or full model scores.
(2)~Dichotomization of these residuals or scores.
(3)~Categorization of possible split variables.
Apart from these three main factors further aspects such as the approximation of the null
distribution (conditional vs.\ unconditional) or the type of test statistic (maximally 
selected vs.\ sum of squares) will be considered as well.
For this purpose, a unifying framework for testing strategies in unbiased model-based tree 
algorithms is presented such that each of the three strategies in GUIDE, CTree, and MOB can be 
obtained by a specific combination of the available building blocks. This allows to systematically
vary the building blocks and assess the power of the resulting inference procedure. Moreover,
it is investigated whether the performance of the inference impacts the performance of the
trees differently under pre- vs.\ post-pruning.

In many of the considered scenarios the choice of goodness-of-fit measure heavily influences
the performance of testing strategies. In particular, using model scores leads to
overall clearly better results than employing residuals only. Moreover, the original
values of the goodness-of-fit measure are preferred over dichotomized versions of them.
Also regarding the effects of categorizing possible split variables and the selection
of a pruning strategy clear recommendations can be given based on the presented results.

@@

@@

>>>>

%\fixme{``Why CTree, GUIDE, and MOB?''}
%
\fixme{Focus on testing strategies}

Therefore, \hlc{instead of comparing the complete algorithms,} in this paper the focus will be put on that first step of tree algorithms,
i.e., the task of selecting the best split variable in a given (sub)sample. 
In particular, the approach from the GUIDE algorithm (which essentially encompasses QUEST)
is compared to the one of the CTree and MOB algorithms by investigating the building blocks of 
their testing strategies in which they differ:
(1)~Variation of the goodness-of-fit measure for the response: using residuals or full model scores.
(2)~Dichotomization of these residuals or scores.
(3)~Categorization of possible split variables.
Apart from these three main factors further aspects such as the approximation of the null
distribution (conditional vs.\ unconditional) or the type of test statistic (maximally 
selected vs.\ sum of squares) will be considered as well.
For this purpose, a unifying framework for testing strategies in unbiased model-based tree 
algorithms is presented such that each of the three strategies in GUIDE, CTree, and MOB can be 
obtained by a specific combination of the available building blocks. This allows to systematically
vary the building blocks and assess the power of the resulting inference procedure. Moreover,
it is investigated whether the performance of the inference impacts the performance of the
trees differently under pre- vs.\ post-pruning.

%% FIXME
\fixme{Provided theory, finite samples}

\hlc{As the underlying theory of the different testing strategies and the resulting properties and skills but also limitations are already discussed in previous literature} \citep[see, e.g.,][]{Ploberger+Kraemer:1992, Hothorn+Hornik+Zeileis:2006, Zeileis+Hothorn+Hornik:2008, Loh:2002}, \hlc{the presented simulation study aims at investigating the corresponding consequences in finite samples.}
In many of the considered scenarios the choice of goodness-of-fit measure heavily influences
the performance of testing strategies. In particular, using model scores leads to
overall clearly better results than employing residuals only. Moreover, the original
values of the goodness-of-fit measure are preferred over dichotomized versions of them.
Also regarding the effects of categorizing possible split variables and the selection
of a pruning strategy clear recommendations can be given based on the presented results.

@@

\fixme{again, focus on testing strategies}

\hlc{For both scenarios the employed testing framework and the considered evaluations allow for a direct comparison of the testing strategies only opposed to the complete tree algorithms.}

@@

\fixme{Outlook: non-tree structure}

\hlc{To exploit the flexibility of tree models future investigations could also include situations where the underlying data generating process does not follow a (clear) tree structure. In this case, assessing the effects of different building blocks in the testing strategies should in general show similar results, however, they might vary depending on how well the true functional relationships can be approximated by a tree structure.}

\fixme{Reference on Fu and Simonoff (2015)}

As \cite{Fu+Simonoff:2015} \hlc{show, unbiased recursive partitioning algorithms can have a bias on their own. In particular, given that there are two possible split variables showing the same significant association to the (transformed) response the algorithms tend to select the one providing less possible split points. Therefore, this could be an additional specific setting in which the power of testing strategies and their building blocks could be compared.}
